\chapter{Evaluation}
\label{ch:evaluation}

The goal of this thesis can be summarized as implementing the \ac{SCITE} algorithm with the same solution quality as the reference implementation, but faster. We describe our design choices in the previous chapter \ref{ch:design}, but we also needed to verify that these choices lead to this goal. This chapter describes how we tested the solution quality of \ac{ffSCITE} using the \ac{TOST} procedure \cite{schuirmann1987comparison} and evaluated its performance using the synthesis report, benchmarking, and profiling. 

Before we can discuss the features of the final build of \ac{ffSCITE}, we need to address a general issue: In section \ref{sec:encoding}, we have decided to limit the maximal number of cells and genes that are processable by the compiled design to an arbitrary number. Therefore, we need to decide on a number of cells and genes for our final design. Our final build of \ac{ffSCITE} supports 64 cells and 63 genes, which is equivalent to mutation trees with 64 nodes. This covers all but one of the example datasets that is bundled with the source code of \ac{SCITE} and compiling the design for 128 cells and 127 genes leads to internal compiler errors we were not able to resolve in time. Lastly, targeting 64 cells and 63 genes also leads to reasonable resource usage which we will discuss later. Therefore, we decided to continue with this input size.

In this chapter, we will first discuss the properties that can be evaluated using the final synthesis report, namely the hardware usage and loop characteristics. We are then using this information to predict the performance of the design and benchmark both \ac{SCITE} and \ac{ffSCITE} to verify this prediction and to evaluate the throughput difference between them. Lastly, we evaluate the solution quality of \ac{ffSCITE}, for which we have to introduce and execute a statistical test. The most important results of these sections are listed in table \ref{tab:quickfacts} for quick look-up.

\begin{table}
    \centering
    \begin{tabular}{r|l}
        \textbf{Metric}                     & \textbf{Value} \\
        \hline
        Max. no. of cells                   & 64 \\
        Max. no. of genes                   & 63 \\
        \hline
        Clock frequency                     & 295.83 MHz \\
        Macro-pipeline capacity             & 6 states \\
        Device utilization (Kernels + DSP)  & 69\% \acsp{LUT} \\
        Predicted throughput                & 577.79 ksteps/s \\
        Measured throughput                 & 566.72 ksteps/s \\
        Maximum speedup                     & 8.63 \\
        \hline
        \ac{FPGA}                           & Intel Stratix 10 GX 2800 \\
        Board                               & Bittware 520N \\
        oneAPI/SYCL version                 & 2022.2.0 Build 133.4 \\
        Quartus version                     & 20.4.0 Build 72 Pro \\
        BSP version                         & 20.4.0\_hpc \\
    \end{tabular}
    \caption{Quick facts about the final \ac{ffSCITE} build.}
    \label{tab:quickfacts}
\end{table}

\input{chapters/evaluation/hardware.tex}
\input{chapters/evaluation/loops.tex}
\input{chapters/evaluation/runtime.tex}
\input{chapters/evaluation/quality.tex}